---
title: Are Capital Cities More Prone to Violent Riots?
author:
  - name: Sonika Mahat
    email: sm6358a@american.edu
    main: true
address: Data Analysis (SIS 750), American University
main_findings:
-  Capital cities report more violent riots, but the difference is not statistically significant.
-  Security forces, demonstrators, students, and protesters are the main actors in capital riots.
output: 
  posterdown::posterdown_betterland:
    self_contained: false
    pandoc_args: --mathjax
    highlight: haddock
    number_sections: false
link-citations: true
bibliography: packages.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = 'asis',
                      echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "50%")
options(knitr.table.format = "html") 
```

# Introduction  

Urban unrest is growing around the world, and capital cities are often at the center of it. Since they hold political power and national institutions, they can be more likely to face conflict. One serious form of unrest is organized violent riots, where violence is used in a planned way by political or identity groups.

This poster looks at two things: Are capital cities more likely to face violent riots than other cities? And who are the main actors involved in these riots? By comparing cities across the world, this research helps us understand how a city’s political role shapes both where and who is involved in urban violence.

# Objectives

This research looks at whether capital cities are more likely to face organized violent riots than non-capitals. Using global city-level data, it explores patterns between political power and urban unrest.

The goal is to better understand how cities can prepare for conflict and make smarter policy decisions around urban safety.

# Data and Methods

This study uses the Urban Social Disorder (USD) 3.0 dataset, which tracks events of political unrest in over 180 cities from 1960 to 2014 (Bahgat et al., 2022). Each event includes details like the type of unrest, the city it happened in, and whether the city is a capital.

I focus on events labeled as organized violent riots (PTYPE = 50), which are planned violent actions by political or identity groups. Using the CAPITAL variable, I compare how often these riots happen in capital cities versus non-capitals.

To see if the difference is meaningful, I use a chi-squared test. I also look at who the main actors are in these riots, especially in capital cities.


# Analysis and Result

The bar chart on the left compares the number of organized violent riots in capital and non-capital cities. Although capital cities show a much higher number of riot events, the chi-squared test result (p = 0.63) suggests that this difference is not statistically significant. This means that, based on this dataset, we cannot say that capital cities are more likely to face such riots purely because they are capitals.

On the right, the chart shows the main actors involved in organized violent riots in capital cities. The most common actors are security forces, demonstrators, and students, followed by protesters. This highlights that both state and non-state groups play key roles in these urban conflicts.

```{r results-analysis, fig.width=10, fig.height=5,out.width="80%", echo=FALSE, message=FALSE, warning=FALSE}
# Load required packages
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(forcats)
library(stringr)
library(patchwork)

events <- read_excel("events.xlsx")
cities <- read_excel("cities.xlsx")

# Merge datasets using CITY_ID
merged_data <- events %>%
  left_join(select(cities, CITY_ID, CAPITAL), by = "CITY_ID")

# Part 1: Capital vs Non-Capital Riot Counts ---------------------

riot_data <- merged_data %>%
  filter(PTYPE == 50)

riot_counts <- riot_data %>%
  count(CAPITAL, name = "Number_of_Riots")

# Plot 1
plot1 <- ggplot(riot_counts, aes(x = CAPITAL, y = Number_of_Riots, fill = CAPITAL)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Organized Violent Riots by Capital Status",
       x = "Capital City",
       y = "Number of Riots") +
  scale_fill_manual(values = c("Yes" = "#00088B", "No" = "#00BFFF")) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 12)
)

# Chi-squared test: Are riots more likely in capital cities?
# Create a contingency table: capital vs. riot event
merged_data <- merged_data %>%
  mutate(riot_event = ifelse(PTYPE == 50, "Riot", "Other"))

table_data <- table(merged_data$CAPITAL, merged_data$riot_event)
chi_result <- chisq.test(table_data)

print(chi_result)
# Part 2: Top Actors in Capital City Riots -----------------------

capital_riots <- merged_data %>%
  filter(PTYPE == 50, CAPITAL == "Yes")

actor_clean <- capital_riots %>%
  select(ACTOR1, ACTOR2, ACTOR3) %>%
  pivot_longer(cols = everything(), names_to = "actor_type", values_to = "actor") %>%
  filter(!is.na(actor), actor != "99") %>%
  mutate(actor = str_to_title(actor),
         actor = case_when(
           str_detect(actor, "Student") ~ "Students",
           str_detect(actor, "Demonstrator") ~ "Demonstrators",
           str_detect(actor, "Protester") ~ "Protesters",
           str_detect(actor, "Police|Security Force") ~ "Security Forces",
           TRUE ~ actor
         )) %>%
  count(actor, sort = TRUE)

# Define which actors to keep
top_actor_names <- c("Students", "Demonstrators", "Security Forces", "Protesters")

# Define custom colors
custom_colors <- c(
  "Students" = "#27408B",
  "Demonstrators" = "#1874CD",
  "Security Forces" = "#00BFFF",
  "Protesters" = "#0000FF"
)

# Filter and reorder only top selected actors
top_cleaned <- actor_clean %>%
  filter(actor %in% top_actor_names) %>%
  mutate(actor = fct_reorder(actor, n))

# Plot
plot2 <- ggplot(top_cleaned, aes(x = actor, y = n, fill = actor)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Top Actors in Capital City Riots",
    x = "Actor",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 12)
  )

# Combine plots
plot1 + plot2
```

# Conclusion

In conclusion, the analysis shows that while capital cities report more organized violent riots, the difference with non-capitals is not statistically significant. The main actors in these riots are security forces, demonstrators, students, and protesters, pointing to recurring tensions between the state and civic groups in capital cities.

# References

Thomson, Henry, Karim Bahgat, Henrik Urdal & Halvard Buhaug 2022. "Urban Social Disorder 3.0: A global, city-level event dataset of political mobilization and disorder." Journal of Peace Research, forthcoming
